---
output:
  html_document: default
  pdf_document: default
  word_document: default
editor_options: 
  chunk_output_type: console
---
### Data Description and Understanding
```{r, include=FALSE}
library(ggplot2)
library(dplyr)
library(naijR)
library(skimr)
library(gmodels)
project_data<- read.csv("C:/Users/User/Desktop/final project/encoded further surv 13 vars.csv")
head(project_data)
```
```{r, include=FALSE}
## lemme try as.factor
project_data$status <- as.factor(project_data$status)
project_data$gender <- as.factor(project_data$gender)
project_data$TOSecSch <- as.factor(project_data$TOSecSch)
project_data$livingINsch <- as.factor(project_data$livingINsch)
project_data$familystructure <- as.factor(project_data$familystructure)
project_data$Deptatsecsch <- as.factor(project_data$Deptatsecsch)
project_data$base <- as.factor(project_data$base)
project_data$sponsor <- as.factor(project_data$sponsor)
project_data$zone <- as.factor(project_data$zone)
project_data$Fathersedu <- as.factor(project_data$Fathersedu)
project_data$Mothersedu <- as.factor(project_data$Mothersedu)
project_data$Educationisthebestlegacy <- as.factor(project_data$Educationisthebestlegacy)
project_data$Schoolisscam <- as.factor(project_data$Schoolisscam)
project_data$opinion<- as.factor(project_data$opinion)
project_data$olevel <- as.factor(project_data$olevel)

```
```{r, echo=FALSE}
skim(project_data)
```
summary of the categorical variables coming through later  

summary of the survivval time, that is time till admission of which 50% is the median survival time   
```{r,echoo = FALSE}
quantile(project_data$survivaltime)
```
### The Contingency Tables  


# Data Presentation and Analysis  
## Non-Parametric Analysis  
```{r, include=FALSE}
library(survival)
library(survminer)
library(survMisc)
library(dplyr)
library(writexl)
#presenting the data from excel 
project_data<- read.csv("C:/Users/User/Desktop/final project/encoded further surv 13 vars.csv")
#attach so you dont have to stress through $ no more
attach(project_data)

```
The **covariates** was reduced from 13 to 6 based on relative importance  
The **covariates** selected are:  
1.  gender  
2.  zone  
3.  tosecsch  
4.  opinion  
5.  livinginsch  
6.  olevelattempts  

```{r, include= FALSE}
##define the variables
time<- project_data$survivaltime
status <- project_data$status
covariates<- cbind(gender,TOSecSch,livingINsch, zone, opinion, olevel)

```
###distribution of responsesacross zones in Nigeria
```{r,echo=FALSE,warning=FALSE}

#create a list of all states by their respective gpz
gpz<-c("ne","nc","nw","ss","se","sw")

statelist<- 
  lapply(gpz,function(x){
    data.frame(State = states(gpz = x), GPZ = toupper(x))
  })
gpz_df<- Reduce(rbind, statelist)
gpz_df<- rbind(gpz_df, c("Federal Capital Territory","FCT"))
map_ng(data = gpz_df,x = GPZ, 
       title = "regional distribution of responses",col = "YlOrBr",
       leg.title = "legend")

```


NC   | 82    
-----|-------
NE   | 3     
-----|-------
NW   | 7     
-----|-------
SE   | 11    
-----|-------
SS   | 12    
-----|-------
SW   | 92    
-----|-------

## kaplan meier non-parametric analysis  
### Overall kaplan-meier estimate and chart 
```{r, echo = FALSE, warning=FALSE}
kmsurv<- survfit(Surv(time, status) ~ 1, data = project_data)
summary(kmsurv)
surv_median(kmsurv)
##****isnt this the table they call life table in S.A
ggsurvplot(kmsurv, fun = "pct", color = "black",
           surv.median.line ="hv",
           conf.int = TRUE,
           conf.int.fill = 'red',
           title ='kaplan-meier survivor curve for the
     "time till first tertiary admission" dataset', xlim = c(0,30),
           font.title = c(16, "bold", "black")
           )

```

## Non-Parametric Analysis by Covariates 
### Hypotheses and log-rank tests  
After charting the kaplan-meier curves, we would like to know if the estimated survivor functions are statistically equivalent.  
first thing first, we set up hypothesis, null(H0) and alternate hypothesis (H1) where H0 is our claim as the researcher and it says " _there is no statistical difference in survival functions (curve) between the levels of covariate_"while the H1 is trying to debunk our claim saying "_there is difference in at least one pair of survival functions_"  
The log-rank test is a large-sample-chi-square test that provides an overall comparison of of the KM curves being compared   
log-rank statistics = $\frac{(O_{2}-E_{2})2}{var(O_{2}-E_{2})}$  

### kmsurvgender<- kaplan-meier estimated curve for gender  
```{r, echo=FALSE}
kmsurvgender<- survfit(Surv(time, status) ~ gender, data = project_data )
summary(kmsurvgender)
surv_median(kmsurvgender)##compare median times of levels
ggsurvplot(kmsurvgender, fun = "pct", linetype = c(1,5), 
           surv.median.line ="hv",
           risk.table = "absolute",
           legend.title = "gender", legend.labs = c("male", "female"),
           title ='Estimated Survivor Curve for Gender', xlim = c(0,30),
           font.title = c(16, "bold", "blue"))
```

### __hypothesis testing__  
H0: there is no difference in survival functions (curve) between the two levels of gender  
H0: s(t_male) = s(t_female)   
H1: there is difference in at least one pair of survival function  
H1: s(t_male) != s(t_female)   

### log rank test  
```{r, echo = FALSE}
survdiff(Surv(time, status) ~gender,rho= 0)
```
### Interpretation of the test
with the log-rank value of 0.6 and a p-value=_0.4_, thus the p-value indicates that the null hypothesis could not be rejected as it follows the criteria for rejection which says __if p-value < 0.05, reject null hypothesis and if otherwise, do not reject H0__  
there was not enough evidence to reject the null hypothesis, therefore the survivor function of male is equal to that of female   
### kmsurv2<- km estimated curve for livinginsch  
```{r,echo=FALSE}
kmsurv2<- survfit(Surv(time, status) ~ livingINsch, data = project_data )
summary(kmsurv2)
ggsurvplot(kmsurv2, fun = "pct", linetype = c(1,5), 
           surv.median.line ="hv",
           risk.table = "absolute",
           legend.title = "livinginsch", legend.labs = c("day", "boarder"),
           title ='Estimated Survivor Curve for Living in School Type', xlim = c(0,30))
surv_median(kmsurv2)##compare median times of levels
```

### Hypothesis
H0: s(t_day) = s(t_boarder)  
H1: s(t_day) != s(t_boarder)  
### log_rank test for living in schol  
```{r, echo = FALSE}
survdiff(Surv(time, status) ~ livingINsch,rho= 0)

```

### Interpretation of the test
with the log-rank value of 5.6 and a p-value=0.02, thus the p-value indicates that the null hypothesis should be rejected based on the criteria: __if p-value < 0.05, reject null hypothesis and if otherwise, do not reject H0__  
there was not enough evidence to reject the null hypothesis.we can therefore conclude that the day and boarder have significantly different KM survival curves    
### kmsurvtos<- km estimated curve for type of ownership of secondary school attended   
```{r echo=FALSE}
kmsurvtos<- survfit(Surv(time, status) ~ TOSecSch, data = project_data )
summary(kmsurvtos)
surv_median(kmsurvtos)##compare median times of levels
ggsurvplot(kmsurvtos, fun = "pct", linetype = c(1,5,9), 
           surv.median.line ="hv",
           risk.table = "absolute",
           legend.title = "Type of Ownership of Secondary School",
           legend.labs = c("federal govt. owned", "state govt. owned", "private"),
           title ='Estimated Survivor Curve for type of ownership of school', xlim = c(0,30))
```

### Hypothesis to test the survivor function of type of ownership of secondary school   
H0: s(t_fed) = s(t_state) = s(t_private)
H1: s(t_fed) != s(t_state) != s(t_private)
#log_rank test for ownership of secondary school
```{r, echo=FALSE}
survdiff(Surv(time, status) ~TOSecSch,rho= 0)
```
### Interpretation of the test  
Since the p-value of the log-rank statistics (0.02) is less than the 0.05 level of significance, we reject H~0~ at 5%level of significance.  
Hence, we concludethat at least one of the groups have a different KM survival curve 

### kmsurv_zone<- km estimated curve for zone  
```{r, echo=FALSE}

kmsurvzone<- survfit(Surv(time, status) ~ zone, data = project_data )
summary(kmsurvzone)
surv_median(kmsurvzone)##compare median times of levels
ggsurvplot(kmsurvzone, fun = "pct", linetype = c(1,5,6,7,8,3), 
           surv.median.line ="hv",
           legend.title = "zone", legend.labs = c("ne","nc","nw","ss","se","sw"),
           title ='Estimated Survivor Curve for Geo-political zone', xlim = c(0,30))
```
### Hypothesis
H0: s(t_NC) = s(t_NE) = s(t_NW) = s(t_SS) = s(t_SW) = s(t_SS)  
H1: s(t_NC) != s(t_NE) != s(t_NW) != s(t_SS) != s(t_SW) != s(t_SS)  

### log_rank test for zone  
```{r,echo=FALSE}
survdiff(Surv(time, status) ~zone,rho= 0)
```
### interpretaton  
log rank test value of 4.7 and a p value of 0.4 says we reject H~0~ as the p-value is less than the alpha value.  
conclusion: we then conclude that there is at least a pair of zones whose KM curves that are different from each other  

### kmsurvopinion<- km estimated curve for opinion  
```{r,echo=FALSE}
kmsurvopinion<- survfit(Surv(time, status) ~ opinion, data = project_data )
summary(kmsurvopinion) 
ggsurvplot(kmsurvopinion, fun = "pct", linetype = c(1,5,4,6,8), 
           surv.median.line ="hv",
           risk.table = "absolute",
           legend.title = "opinion",
           legend.labs = c("strong +ve attitude to edu and sch", 
                           "+ve attitude to edu and sch",
                           "neutral",
                           "-ve attitude to edu and sch",
                           "strong -ve attitude to edu and sch"),
           title ='Estimated Survivor Curve for opinion', xlim = c(0,30),
           font.title = c(16, "bold", "blue"))
```
**compare median times of levels**    
```{r, echo=FALSE}
surv_median(kmsurvopinion)
```
### Hypothesis for opinion   
H0: s(t_s+) = s(t_+) = s(t_neutral) = s(t_-) = s(t_s-)  
H1: s(t_s+) != s(t_+) != s(t_neutral) != s(t_-) != s(t_s-)  
# log_rank test for opinion  
```{r,echo = FALSE}
survdiff(Surv(time, status) ~opinion,rho= 0)

```
### Interpretation of the test  
the p-value of the log-rank test at 5% level of significance is 0.6 which is greater than the alpha. Hence, H~0~ is not   rejected and we conclude that the groups of opinion do not have different KM survivor curves   
 
### kmsurvexam<- km estimated curve for no of times respondents tried 0level examination   
```{r,echo=FALSE}
kmsurvexam<- survfit(Surv(time, status) ~ olevel, data = project_data )
summary(kmsurvexam) 
ggsurvplot(kmsurvexam, fun = "pct", linetype = c(1,5), 
           surv.median.line ="hv",
           risk.table = "absolute",
           legend.title = "gender", legend.labs = c("once", "more than once"),
           title ='Estimated Survivor Curve for opinion', xlim = c(0,30),
           font.title = c(16, "bold", "blue"))
```

### compare median times of levels  
```{r,echo=FALSE}
surv_median(kmsurvexam)
```
### hypothesis  
H0: s(t_once) = s(t_more)  
H1: s(t_once) = s(t_more)   

### log_rank test for olevel  
```{r, echo=FALSE}
survdiff(Surv(time, status) ~olevel,rho= 0)

```
# Interpretation and Conclusion
with the log-rank value of 3.5 and a p-value=0.06 which is greater than the 5% level of significance, thus the p-value indicates that H~0~ should not be rejected based on the criteria: __if p-value < 0.05, reject null hypothesis and if otherwise, do not reject H0__    
**Conclusion** there was not enough evidence to reject the null hypothesis. We can therefore conclude that the two survival groups do not have statistically different KM survival curves  

### Nelson-Aalen Cummulative Hazard 
below is the Nelson-Aalen cummulative hazard estimate  
```{r,echo=FALSE}
ggsurvplot(kmsurv, fun = "cumhaz")

```

dummy variable encoding
```{r, echo=FALSE}
##dummy encoding

##dummy for zone where nc is the reference variable
sw<-ifelse(zone=='6',1,0)
ss<-ifelse(zone=='5',1,0)
se<-ifelse(zone=='4',1,0)
nw<-ifelse(zone=='3',1,0)
ne<-ifelse(zone=='2',1,0)

##dummy for gender where male is the reference variable
female<-ifelse(gender=='2',1,0)

##dummy for type of secondary school where fed is the reference variable
state<-ifelse(TOSecSch=='2',1,0)
private<-ifelse(TOSecSch=='3',1,0)

##dummy for livinginsch where boarder is the reference variable
day<-ifelse(livingINsch=='1',1,0)

##dummy for opinion where strong negative is the reference variable
strong_positive<-ifelse(opinion=='1',1,0)
positive<-ifelse(opinion=='2',1,0)
neutral<-ifelse(opinion=='3',1,0)
negative<-ifelse(opinion=='4',1,0)

##dummy for olevel exam where more is the reference variable
once<-ifelse(olevel=='0',1,0)

dummy_covariates<- cbind(sw,ss,se,nw,ne,
                    female,
                    state,private,
                    day,
                    strong_positive,positive,neutral,negative,
                    once)

head(dummy_covariates)
```

```{r,echo = FALSE}
cfit1<- coxph(Surv(time, status)~ gender + TOSecSch +livingINsch + zone                                           + opinion +olevel,data= project_data)
cfit1
```
```{r, echo = FALSE}
cfit<- coxph(Surv(time, status)~ sw+ss+se+nw+ne+female+state+private+day+ strong_positive +positive + neutral +negative +once,
             data = project_data, method = "breslow")

cfit
```



